Reasoning Graphically about Quantum Computation

### 2008-02-27 papers/reasoning-graphically/paper.tex
re-organised sections, !-box graphs section written... more stuff...
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1378 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

The representation of Compact Closed Categories as Graphs, discuseed
in the previous section, is too restrictive for automated reasoning
about quantum computation. In particular, pictures such the Spider
Law, in Figure~\ref{spider-law-fig} are frequently needed. Such
pictures represent an infinite family of concrete graphs. In this
section we describe an extension to the representation that captures
an infinite family of graphs as a single finite graph. The extension,
which we call \ref{graph patterns}, adds two extra kinds of nodes to
the graph. Namely, {\em banged boxes} which informally represent an
arbitrary number of copies of a node (where the node itself can
contain a graph of the appropriate type), and {\em variable nodes}
which informally represent some, as yet unkown, node in a graph. We
also introduce a notion to replace graph composition, called graph
plugging that allows graph patterns to put together to form larger
graph patterns.

%Notably, this
%allows us to represent laws about quantum computation, such as the
%Spider Law. Furthermore, this representation is still compositional in
%the sense that we can plug graphs together.

%\begin{center}
%  \scalebox{1.0}{\includegraphics{images/spider_lhs}}
%\end{center}

\subsection{Variable Nodes}

In graph patterns, variable nodes generalise the concept of a boundary
node. Variable nodes, unlike boundary-nodes in the concrete graph, are
not classified into domain and co-domain. The intuition of a variable
node is that it can be replaced by any subgraph of the right
type. Thus, unlike nodes in conrete a graph, variable nodes do not
define an ordering on the edges that are incident. Instead, the edges
that touch a variable node are simply a set. The type of a variable
nodes is thus defined as a multiset rather than a tensor. Any tensor
type can be lifted to a multiset type in the obvious manner, which we
denote by the operation {\tt mset\_of\_tensor}. For example, {\tt
  mset\_of\_tensor($a \tensor b \tensor a$) = $\{(a,2), (b,1)\}$}. We
say that a tensor-type, $T$, is an instance of a multiset type, $M$,
when it's lifting into a multiset is a super-set of $M$: $ M \subseteq
\mathtt{mset\_of\_tensor}(T) \Longrightarrow T \in M$. The intuition
here is that an instance of a variable node is allowed to have as many
extra edges as it likes as long as it at least has the ones mentioned
in the original graph.

@@

\subsection{!-Boxes}

@@

For the purposes of this paper, the underlying graph representation of
!-boxed graphs is variable-node graphs. The set of variable-node
graphs for a graph $G$ with !-boxes, is denoted by $\binterp{G}$. We
now define the semantics for !-boxed graphs, then show how to extend
an the embedding function which corresponds to matching as well as the
plugging operation for !-boxes.

\subsubsection{!-Box Semantics.}

We give a formal semantics to !-boxing a representation of graphs by
defining the a !-box graph in terms of a set of graphs in the
underlying representation. We denote this interpretation of a graph
$G$ by $\binterp{G}$ and say that mebemrs of $\binterp{G}$ are
instances of $G$. The definition of $\binterp{G}$ involves the
following three operations\footnote{These correspond closely to the
  linear logic rules for weakening, contraction and deriliction.} on a
!-box and a graph that contains it:

\begin{definition}
  \emph{copy} copies a box in a graph to produce a new graph with two
  copies of the boxed subgraph, with a second box around the newly
  copied nodes. Any edges incident on both a node in the !-box and
  outside it get copied so that there the old edge remains and the new
  edge goes from the outer node to the new copy of the internal node.
\end{definition}

@@

\begin{definition}
  \emph{kill} removes from the graph all nodes in the !-box set as
  well as any incident edges to nodes in the box.
\end{definition}

\noindent The semantics of a !-boxed graph is then defined as those
members of the set closed by these operation that do have any
!-boxes. Observe that an instance of a !-box graph is defined by
pairing each !-box with the natural number that defines how many
copies are made of it. Thus the $\binterp{G}$ is isomorphic to the set
of $k$-tuples of natural numbers, where $k$ is the number of !-boxes.

\subsubsection{!-Box Matching.}

Matching for !-box graphs is defined in terms of an underlying
matching operation. A !-box graph $G$ matches another one, $H$, iff
$\binterp{G} \supseteq \binterp{H}$. Matching can be given a more
constructive flavour by extending the open-embeddings of variable-node
graphs to !-boxes.

There is a !-open embedding of $G$ into $H$ if

\subsubsection{!-Box Plugging.}

@@

>>>>

\section{Graphs with Variable Nodes}

Variable nodes generalise the concept of a boundary nodes. Unlike
boundary-nodes in the concrete graph, they are not classified into
domain and co-domain. The intuition of a variable node is that it can
be replaced by any subgraph of the right type. Thus, unlike nodes in
conrete a graph, variable nodes do not define an ordering on the edges
that are incident. Instead, the edges that touch a variable node are
simply a set. The type of a variable nodes is thus defined as a
multiset rather than a tensor. Any tensor type can be lifted to a
multiset type in the obvious manner, which we denote by the operation
{\tt mset\_of\_tensor}. For example, {\tt mset\_of\_tensor($a \tensor
  b \tensor a$) = $\{(a,2), (b,1)\}$}. We say that a tensor-type, $T$,
is an instance of a multiset type, $M$, when it's lifting into a
multiset is a super-set of $M$: $ M \subseteq
\mathtt{mset\_of\_tensor}(T) \Longrightarrow T \in M$. The intuition
here is that an instance of a variable node is allowed to have as many
extra edges as it likes as long as it at least has the ones mentioned
in the original graph.

@@

\section{!-Boxes}

@@

For the purposes of this paper, the underlying graph representation of
!-boxed graphs is variable-node graphs. The set of variable-node
graphs for a graph $G$ with !-boxes, is denoted by $\binterp{G}$. We
now define the concept of !-box matching which we use to provide a
semantics for the !-boxed graphs.

\subsection{!-Matching.}

The set of !-box graphs matched by a !-box graph is defined as the set
closed by the following operations:

\begin{definition}
  \emph{copy} copies a !-box in a graph to produce a new graph with
  two copies of the boxed subgraph, with a second box around the newly
  copied nodes. Any edges between a node inside the !-box, $n$ and one
  outside it $m$ get copied so that there is a new edge from $m$ to
  the new copy of $n$.
\end{definition}

@@

\begin{definition}
  \emph{kill} removes from the graph all nodes in the !-box set as
  well as any incident edges to those nodes.
\end{definition}

\begin{definition}
  \emph{merge} combines two !-boxes, $B_1$ and $B_2$ into a single
  larger !-box $B_1 \union B_2$.
\end{definition}

Like the notation for variable-node graphs, we use $G \leq H$ to
denote that $G$ matches $H$.

\subsection{!-Box Semantics.}

We give a formal semantics to !-boxing a representation of graphs by
defining the !-box graph in terms of a set of graphs in the
underlying representation. We denote the interpretation of a graph $G$
by $\binterp{G}$ and say that mebemrs of $\binterp{G}$ are instances
of $G$.

\begin{definition}
  \emph{Interpretation of !-box Graphs}: $\binterp{G}$ is the subset
  of graphs matched by the !-box graph that have no !-boxes.
\end{definition}

Observe that an instance of a !-box graph is defined by pairing each
!-box with the natural number that defines how many copies are made of
it. Thus the $\binterp{G}$ is isomorphic to the set of $k$-tuples of
natural numbers, where $k$ is the number of !-boxes.

\begin{theorem}
  \emph{!-Matching respects !-box semantics}: $G \leq H
  \Leftrightarrow \binterp{G} \supseteq \binterp{H}$. The proof is a
  simple consequence from the defintion of $\binterp{G}$ being a
  subset of the graphs that match $G$.
\end{theorem}

The representation of Compact Closed Categories as Graphs, discuseed
in the previous section~\S\ref{graph-repr-comp}, is too restrictive
for reasoning about quantum computation. In particular, pictures such
the Spider Law, in Figure~\ref{spider-law-fig} are frequently
needed. Such pictures apply to an infinite family of concrete
graphs. In this section we combine {\em !-boxes} with {\em variable
  nodes} to form a representation of graphs that allows us to finitely
express certain infinite families of concrete graphs. We call this
extension \ref{graph patterns}. Using this representation Spider Law
can now be represented by the equation on graph paterns shown in
Figure~\ref{graph-pattern-spider-law-fig}. Graph matching for pattens
graphs is defined as is a suitable extension to plugging. This
provides the foundational machinery for a rewriting system, based on
pattern graphs, for reasoning about quantum computation.

@@

We then extend the plugging
operation on variable-node graphs to graphs with !-boxes.

\subsection{Plugging}
